Pervasive duplication of tumor suppressor genes preceded parallel evolution

### 2020-06-03 paper_PLOS/paper_PLOS_draft.Rmd
Refactored figures and  tables, fixed Figure 6, broke Figure 4&7.

---
title: Pervasive duplication of tumor suppressor genes preceded parallel evolution
  of large bodied Atlantogenatans
author:
- name: Juan Manuel Vazquez
  affiliation: "University of Chicago"
- affiliation: "SUNY Buffalo"
  name: Vincent J Lynch
address:
- address: Department of Human Genetics, 920 East 58th St, Chicago, IL, 60637
  code: "University of Chicago"
- address: Department of Biological Sciences, 551 Cooke Hall, Buffalo NY, 14260
  code: "SUNY Buffalo"
output:
  rticles::elsevier_article:
    includes:
      in_header: preamble.tex
  pdf_document: default
  word_document: default
  workflowr::wflow_html: default
csl: plos.csl
editor_options:
  chunk_output_type: console
bibliography: rbhb_paenungulates.bib
abstract: "Cancer is an intrinsic disease of multicellular organisms. Within a species,
  the size of an animal, - correlated with the individual's number of cells - and
  its lifespan - correlated with increasing cellular damage over time - positively
  correlate with the risk any individual has to form tumors. Between species, however,
  we do not observe any correlation between size, lifespan, and cancer, a phenomenon
  that referred to as _Peto's Paradox_. Elephants are a particularly interesting member
  of this class of paradoxical animals, since they are a set of large species deeply
  nested in a clade of smaller species, indicating a recent gain of size. Recent work
  has identified several individual cases of gene duplicates contributing to the increased
  cancer resistance of elephants, which suggests that duplication of tumor suppressor
  genes may play a more general role in mediating Peto's Paradox by increasing cancer
  resistance in large, long-lived species. \nBy using a Reciprocal Best-Hit BLAT search
  approach, we investigated copy numbers of all protein-coding genes in _Atlantogenatan_
  genomes to see if there is any correlation between the copy number of duplicates
  and changes body size along the phylogenetic tree. From an initial set of 18,011
  protein-coding genes in hg38, we identified a median of 13,880 genes in _Atlantogenatan_
  genomes, of which a median of 940 genes are duplicated. We find that, just as body
  size fluctuates throughout _Atlantogenata_, genes involved in tumor suppressor pathways
  are also duplicated throughout the phylogenetic tree. Extant species of elephants,
  however, show active transcription of both canonical and duplicated copies of tummor
  suppressors that duplicated prior to and during their sudden increase in body size,
  suggesting that the duplication of tumor suppressor genes facilitates the evolution
  of increased body size by compensating for the increased cancer risk. \n"

@@

    ```{r Libraries and functions, include = F}
knitr::opts_chunk$set(eval=T, echo=F, message = F, warning = F, # fig.width=6, 
                      fig.retina = 2, fig.align = "left", dpi = 300)

@@

t.bodysize.atlantogenata.withNodes <- st.Atlantogenata %>% 
  as_tibble() %>% 
  filter(!is.na(label)) %>%
  select(
    `Node`=label, 
    `Size (log(g))`=Median, 
    `95% CI (Low)` = `95%CI_Low`, 
    `95% CI (High)` = `95%CI_High`,  
    `Rate (sqrt)`=sqrt.rate
  ) %>%
  mutate_at(vars(-`Node`, ), . %>% as.numeric %>% round(2)) %>% 
  kable(
    ., 
    longtable = TRUE, 
    booktabs = TRUE,
    caption = 'All body size estimates for _Atlantogenata_.'
  )

@@

```{r Figure0Prep}

@@

```
```{r Figure0, fig.cap="Atlantogenatans with sequenced genomes, body sizes, and known lifespans [@PuttickAndThomas2015; @HAGR]",  out.width="6in", fig.asp=1.3}
atlantogenata.tree.size.lifespan
```

@@

One of the major constraints on the evolution of large body sizes in animals is an increased risk of developing cancer. If all cells in all organisms have a similar risk of malignant transformation and equivalent cancer suppression mechanisms, organism with many cells should have a higher prevalence of cancer than organisms with fewer cells.  Consistent with this expectation there is a strong positive correlation between body size and cancer incidence within species, for example, human cancer incidence increases with increasing adult height [@Green2011; @Nunney:20181c2] and cancer incidence is positively correlated with body size in dogs [@Dobson2013; @Dorn1968]. There is no correlation, however, between body size and cancer risk between species. This lack of correlation is often referred to as ‘Peto’s Paradox’ [@Caulin2011; @Leroi2003; @Peto1975]. While it is clear that a resolution to Peto’s Paradox must involve the evolution of enhanced cancer protection alongside increases in body size and lifespan, the specific genetic, molecular, and cellular mechanisms that underlie this resistance have proven elusive. [c.f. @Ashur-Fabian2004; @Seluanov2008; @Gorbunova2012; @Tian2013; @Sulak2015].

Among the challenges for discovering how animals evolved enhanced cancer protection mechanisms is identifying lineages in which large bodied species are nested within species with small body sizes. Afrotherian mammals are generally small-bodied, similarly to the prediced common ancestor of Eutherian mammals. For example, maximum adult weights are ~70g in golden moles, ~120g in tenrecs, ~170g in elephant shrews, ~3kg in hyraxes, and 60kg in aardvarks [@HAGR]. However, while these extant species are relatively small, the fossil evidence demonstrates that their ancestral lineages reached enormous sizes. For example, while extant hyraxes are relatively small, the extinct Titanohyrax is estimated to have weighted up to ~1300kg [@Schwartz1995]. The largest members of Afrotheria, too, are dwarfed by the size of their recent ancestors: extant cows manatees are large bodied (~322-480kg) but are relatively small compared to the extinct Stellar’s sea cow which is estimated to have weight 8000-10000kg [@Scheffer1972]. Similarly African (4,800kg) and Asian elephants (3,200kg) are the largest living elephant species, but are dwarfed by the truly gigantic extinct Proboscideans such as Deinotherium (~132,000kg), Mammut borsoni (110,000kg), and the Asian straight-tusked elephant (~220,000kg), the largest known land mammal [@Larramendi:20151c2]. These large-bodied Afrotherian lineages are nested within small bodied species (Fig. 1) [@OLeary2013a; @Springer2013; @OLeary2013b; @PuttickAndThomas2015], indicating that gigantism independently evolved in hyraxes, sea cows, and elephants (Paenungulates). Thus, Paenungulates are an excellent model system in which to explore the mechanisms that underlie the evolution of large body sizes and augmented cancer resistance.

Although many mechanisms can potentially resolve Peto’s paradox, among the most parsimonious routes to enhanced cancer resistance is through an increased copy number of tumor suppressors. Such an example has been seen in the case of candidate genes such as _TP53_ and _LIF_ [@Abegglen:JAMA2015; @Sulak2015; @Vazquez2018] as well as in studies involving a limited set of candidate genes [@Caulin2015; @Doherty2016]. As these studies focus on _a priori_ gene sets, however, it remains unknown whether this is a general, genome-wide trend in Afrotherian genomes; and whether such a general trend is associated with the recent increases in body size – and therefore expected cancer risk – in these species.

Here, we trace the evolution of body mass and gene copy number variation in Afrotherians in order to investigate whether gene duplications are enriched in large, long-lived species for genes involved in known tumor suppression pathways. Our estimates of the evolution of body mass, similarly to previous studies [@OLeary2013a; @Springer2013; @OLeary2013b; @PuttickAndThomas2015], show that large body masses evolved in a step-wise manner, with major increases in body mass in the Pseudoungulata (17kg), Paenungulata (25kg), Tethytheria (296kg), and Proboscidea (4,100kg) stem-lineages. Furthermore, we see that the ancestral body size increases in Hydracoidia and Sirenia were independent events. To study the evolution of gene copy number, we used a genome-wide Reciprocal Best BLAT Hit (RBBH) method to identify gene duplications in Afrotherian genomes, and used maximum likelihood (treating copy number as a discrete trait) to infer the lineages in which those duplications occurred. We found gene duplications in lineages with increased body mass were enriched in functions related to tumor suppression, including regulation of the cell cycle, DNA damage repair, and regulation of apoptosis. These data suggest that duplication of tumor suppressors played a role in the evolution of large, long-lived in Afrotherians.

@@

We built a time-calibrated supertree of Eutherian mammals by combining the time-calibrated molecular phylogeny of Bininda-Emonds _et al._ [-@Bininda-Emonds2008] with the time-calibrated total evidence Afrotherian phylogeny from Puttick and Thomas [@PuttickAndThomas2015]. While the Bininda-Emonds _et al._ [-@Bininda-Emonds2008] phylogeny includes 1,679 species, only 34 are Afrotherian, and no fossil data are included. The inclusion of fossil data from extinct species is essential to ensure that ancestral state reconstructions of body mass are not biased by only including extant species. This can lead to inaccurate reconstructions, for example, if lineages convergently evolved large body masses from a small bodied ancestor. In contrast, the total evidence Afrotherian phylogeny of Puttick and Thomas [-@PuttickAndThomas2015] includes 77 extant species and fossil data from 39 extinct species. Therefore we replaced the Afrotherian clade in the Bininda-Emonds _et al._ [-@Bininda-Emonds2008] phylogeny with the Afrotherian phylogeny of Puttick and Thomas [-@PuttickAndThomas2015] using Mesquite. Next, we jointly estimated rates of body mass evolution and reconstructed ancestral states using a generalization of the Brownian motion model that relaxes assumptions of neutrality and gradualism by considering increments to evolving characters to be drawn from a heavy-tailed stable distribution (the "Stable Model") [@ElliotAndMooers2014]. The stable model allows for occasional large jumps in traits and has previously been shown to out-perform other models of body mass evolution, including standard Brownian motion models, Ornstein–Uhlenbeck models, early burst maximum likelihood models, and heterogeneous multi-rate models [@ElliotAndMooers2014].

@@

__Reciprocal Best-Hit BLAT:__ We developed a reciprocal best hit BLAT (RBHB) pipeline to identify putative homologs and estimate gene copy numbers across species (__Figure 1A__). The Reciprocal Best Hit (RBH) search strategy is conceptually straightforward: 1) Given a gene of interest $G_A$ in a query genome $A$, one searches a target genome $B$ for all possible matches to $G_A$; 2) For each of these hits, one then performs the reciprocal search in the original query genome to identify the highest-scoring hit; 3) A hit in genome $B$ is defined as a homolog of gene $G_A$ if and only if the original gene $G_A$ is the top reciprocal search hit in genome $A$. We selected BLAT [@blat] as our algorithm of choice, as this algorithm is sensitive to highly simliar (>90% identity) sequences, thus identifying the highest-confidence homologs while minimizing many-to-one mapping problems when searching for multiple genes. RBH performs similar to other more complex methods of orthology prediction, and is particularly good at identifying incomplete genes that may be fragmented in low quality/poor assembled regions of the genome [@AltenhoffAndDessimoz2009; @SalichosAndRokas2011].

__Effective Copy Number By Coverage:__ In lower-quality genomes, many genes are fragmented across multiple scaffolds, which results in BLAT calling multiple hits when in reality there is only one gene. To compensate for this, we came up with a novel statistic, Estimated Copy Number by Coverage (ECNC), which averages the number of times we see each nucleotides of a query sequence in a target genome over the total number of nucleotides of the query sequence found overall in each target genome (Supplementary Figure 1). This allows us to correct for genes that have been fragmented across incomplete genomes, while also taking into account missing sequences from the human query in the target genome. Mathematically, this can be written as:  
$$ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}$$
where $n$ is a given nucleotide in the query, $l$ is the total length of the query, $C_n$ is the number of instances that $n$ is present within a reciprocal best hit, and $bool(C_n)$ is 1 if $C_n > 0$ or 0 if $C_n = 0$.  
```{r RBHB Stats}
n.TranscriptsSearched <- query.binTable %>% filter(!C.ORF, !LOC, !LINC, !HLA, !Histone, !ZincFinger,!Riboprotein, !OlfactoryReceptor, !Uncharacterized, !Putative, !Viral, !Fragment) %>% pull(ID) %>% unique %>% length()
n.GenesSearched <- query.binTable %>% filter(!C.ORF, !LOC, !LINC, !HLA, !Histone, !ZincFinger,!Riboprotein, !OlfactoryReceptor, !Uncharacterized, !Putative, !Viral, !Fragment) %>% pull(Gene) %>% unique %>% length()
genesPerGenome <- geneCopyTable %>% filter(Copy>=1, !is.na(ECNC)) %>% group_by(Genome) %>% summarise(
  Hits=n(), pc.Found=Hits/n.GenesSearched, 
  mean.ECNC = mean(ECNC), q1.ECNC = quantile(ECNC, 1/4), median.ECNC = median(ECNC), q3.ECNC = quantile(ECNC, 3/4), max.ECNC = max(ECNC), sd.ECNC = sd(ECNC), 
  mean.Copy = mean(Copy), q1.Copy = quantile(Copy, 1/4), median.Copy = median(Copy), q3.Copy = quantile(Copy, 3/4), max.Copy = max(Copy), sd.Copy = sd(Copy), 
  mean.ECNCPerHit = mean(ECNC/Copy), q1.ECNCPerHit = quantile(ECNC/Copy, 1/4), median.ECNCPerHit = median(ECNC/Copy), q3.ECNCPerHit = quantile(ECNC/Copy, 3/4), max.ECNCPerHit = max(ECNC/Copy), min.ECNCPerHit = min(ECNC/Copy), sd.ECNCPerHit = sd(ECNC/Copy))
duplicatedGenesPerGenome <- geneCopyTable %>% filter(Copy>=1, ECNC>=1.5) %>% group_by(Genome) %>% summarise(Duplicated=n())

@@

After filtering out problematic protein queries (see "Query gene inclusion criteria"), we then used our pipeline (Figure 1A) to search for all copies of our `r n.GenesSearched` query genes in publicly available Afrotherian genomes, including African savannah elephant (_Loxodonta africana_: loxAfr3, loxAfr4, loxAfrC), African forest elephant (_Loxodonta cyclotis_: loxCycF), Asian Elephant (_Elephas maximus_: eleMaxD), Woolly Mammoth (_Mammuthus primigenius_: mamPriV), Colombian mammoth (_Mammuthus columbi_: mamColU), American mastodon (_Mammut americanum_: mamAmeI), Rock Hyrax (_Procavia capensis_: proCap1, proCap2, proCap2_HiC), West Indian Manatee (_Trichechus manatus latirostris_: triManLat1, triManLat1_HiC), Aardvark (_Orycteropus afer_: oryAfe1, oryAfe1_HiC), Lesser Hedgehog Tenrec (_Echinops telfairi_: echTel2), Nine-banded armadillo (_Dasypus novemcinctus_: dasNov3), Hoffman's two-toed sloth (_Choloepus hoffmannii_: choHof1, choHof2, choHof2_HiC), Cape golden mole (_Chrysochloris asiatica_: chrAsi1), and Cape elephant shrew (_Elephantulus edwardii_: eleEdw1). For many of these species, we covered multiple assemblies in order to test the effects of assembly size and quality on our hits.

__Query gene inclusion criteria:__ To assemble our query list, we first removed all unnamed genes from UP000005640. Next, we excluded genes from downstream analyses for which assignment of homology was uncertain, including uncharacterized ORFs (`r query.binTable %>% filter(C.ORF) %>% nrow`), LOC (`r query.binTable %>% filter(LOC) %>% nrow`), HLA genes (`r query.binTable %>% filter(HLA) %>% nrow`), replication dependent histones (`r query.binTable %>% filter(Histone) %>% nrow`), odorant receptors (`r query.binTable %>% filter(OlfactoryReceptor) %>% nrow`), ribosomal proteins (`r query.binTable %>% filter(Riboprotein) %>% nrow`), zinc finger transcription factors (`r query.binTable %>% filter(ZincFinger) %>% nrow`), viral and repetitive-element-associated proteins (`r query.binTable %>% filter(Viral) %>% nrow`) and any protein described as either "Uncharacterized," "Putative," or "Fragment" by UniProt in UP000005640 (`r query.binTable %>% filter(Fragment | Uncharacterized | Putative) %>% nrow`), leaving us with a final set of `r n.TranscriptsSearched` query protein sequences, corresponding to `r n.GenesSearched` genes.

__Duplication gene inclusion criteria:__ In order to condense transcript-level hits into single gene loci, and to resolve many-to-one genome mappings, we removed exons where transcripts from different genes overlapped, and merged overlapping transcripts of the same gene into a single gene locus call. The resulting gene-level copy number table was then combined with the maximum ECNC values observed for each gene in order to call gene duplications. We called a gene duplicated if its copy number was two or more, and if the maximum ECNC value of all the gene transcripts searched was 1.5 or greater; previous studies have shown that incomplete duplications can encode functional genes [@Sulak2015; @Vazquez2018], therefore partial gene duplications were included provided they passed additional inclusion criteria. The ECNC cut off of 1.5 was selected empirically, as this value minimized the number of false positives seen in a test set of genes and genomes. The results of our initial search are summarized in Figure 1B`r warning('TODO: Check if this is still the case')`. Overall, we identified `r median.genesPerGenome` genes across all species, or `r median.genesPerGenome %>% divide_by(n.GenesSearched) %>% multiply_by(100) %>% round(digit=1) %>% paste0(.,"%")` of our starting query genes.

@@

## Evidence for Functionality of Identified Genes

@@

__Transcriptome Assembly:__ For the African Savana Elephant, Asian Elephant, West Indian Manatee, and Nine-Banded Armadillo, we generated _de novo_ transcriptomes using publically-available RNA-sequencing data from NCBI SRA (Supplementary Table 1). We mapped reads to all genomes available for each species, and assembled transcripts using HISAT2 and StringTie, respectively [@HISAT; @StringTie; @Tuxedo]. RNA-sequencing data was not available for Cape Golden Mole, Cape Elephant Shrew, Rock Hyrax, Aardvark, or the Lesser Hedgehog Tenrec.

@@

To determine which pathways were associated with duplicated genes in each species and lineage, we used WEBGESTALT to perform overrepresentation analysis (ORA) of the duplicated gene lists relative to our initial query gene list [@WebGestalt2019]. For the database of pathways used in the analysis, we used Reactome [@Reactome], Wikipathways, [@Wikipathways], and KEGG [@KEGG]. For the ORA, we used FDR for determining significance, and ran the analysis at FDR=0.1, FDR=0.2, FDR=0.3, and FDR=0.5.

## Lifespan Phylogenetic Least-Square Regression and Calculating Estimated Cancer Risk Thoughout Atlantogenata

>>>>

---
title: Pervasive duplication of tumor suppressor genes preceded parallel evolution
  of large bodied Atlantogenatans
author:
- name: Juan Manuel Vazquez
  affiliation: "University of Chicago"
- affiliation: "SUNY Buffalo"
  name: Vincent J Lynch
address:
- address: Department of Human Genetics, 920 East 58th St, Chicago, IL, 60637
  code: "University of Chicago"
- address: Department of Biological Sciences, 551 Cooke Hall, Buffalo NY, 14260
  code: "SUNY Buffalo"
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    number_sections: yes
    includes:
      in_header: preamble.tex
  pdf_document: default
  word_document: default
  workflowr::wflow_html: default
csl: plos.csl
editor_options:
  chunk_output_type: console
bibliography: rbhb_paenungulates.bib
abstract: "Cancer is an intrinsic disease of multicellular organisms. Within a species,
  the size of an animal, - correlated with the individual's number of cells - and
  its lifespan - correlated with increasing cellular damage over time - positively
  correlate with the risk any individual has to form tumors. Between species, however,
  we do not observe any correlation between size, lifespan, and cancer, a phenomenon
  that referred to as _Peto's Paradox_. Elephants are a particularly interesting member
  of this class of paradoxical animals, since they are a set of large species deeply
  nested in a clade of smaller species, indicating a recent gain of size. Recent work
  has identified several individual cases of gene duplicates contributing to the increased
  cancer resistance of elephants, which suggests that duplication of tumor suppressor
  genes may play a more general role in mediating Peto's Paradox by increasing cancer
  resistance in large, long-lived species. \nBy using a Reciprocal Best-Hit BLAT search
  approach, we investigated copy numbers of all protein-coding genes in _Atlantogenatan_
  genomes to see if there is any correlation between the copy number of duplicates
  and changes body size along the phylogenetic tree. From an initial set of 18,011
  protein-coding genes in hg38, we identified a median of 13,880 genes in _Atlantogenatan_
  genomes, of which a median of 940 genes are duplicated. We find that, just as body
  size fluctuates throughout _Atlantogenata_, genes involved in tumor suppressor pathways
  are also duplicated throughout the phylogenetic tree. Extant species of elephants,
  however, show active transcription of both canonical and duplicated copies of tummor
  suppressors that duplicated prior to and during their sudden increase in body size,
  suggesting that the duplication of tumor suppressor genes facilitates the evolution
  of increased body size by compensating for the increased cancer risk. \n"

@@

```{r Libraries and functions, include = F}
knitr::opts_chunk$set(eval=T, echo=F, message = F, warning = F, out.width="6in",
                      fig.retina = 2, fig.align = "left", dpi = 300)

@@

@@

```{r Figure1Prep}

@@

@@

Among the major constraints on the evolution of large body sizes (and long life-spans) in animals is an increased risk of developing cancer. If all cells in all organisms have a similar risk of malignant transformation and equivalent cancer suppression mechanisms, organism with many cells should have a higher prevalence of cancer than organisms with fewer cells, particularly because large and small animals have similar cell sizes (@Savage2007). Consistent with this expectation there is a strong positive correlation between body size and cancer incidence within species, for example, cancer incidence increases with increasing adult height in humans [@Green2011; @Nunney:20181c2] and in dogs [@Dobson2013; @Dorn1968]. There is no correlation, however, between body size and cancer risk _between_ species; this lack of correlation is often referred to as ‘Peto’s Paradox’ [@Caulin2011; @Leroi2003; @Peto1975]. The ultimate resolution to Peto’s Paradox is obvious: large bodied and long-lived species evolved enhanced cancer protection mechanisms. However, identifiying the specific genetic, molecular, and cellular mechanisms that underlie the evolution of augmented cancer protection has been difficult [c.f. @Ashur-Fabian2004; @Seluanov2008; @Gorbunova2012; @Tian2013; @Sulak2015].

Among the challenges for discovering how animals evolved enhanced cancer protection mechanisms is identifying lineages in which large bodied species are nested within species with small body sizes. _Afrotherian_ mammals are generally small-bodied, but also include the largest extant land mammals. For example, maximum adult weights are ~70g in golden moles, ~120g in tenrecs, ~170g in elephant shrews, ~3kg in hyraxes, and 60kg in aardvarks [@HAGR]. While extant hyraxes are relatively small, the extinct Titanohyrax is estimated to have weighted up to ~1300kg [@Schwartz1995]. The largest members of _Afrotheria_, too, are dwarfed by the size of their recent ancestors: extant cows manatees are large bodied (~322-480kg) but are relatively small compared to the extinct Stellar’s sea cow which is estimated to have weight 8000-10000kg [@Scheffer1972]. Similarly African (4,800kg) and Asian elephants (3,200kg) are the largest living elephant species, but are dwarfed by the truly gigantic extinct Proboscideans such as Deinotherium (~132,000kg), Mammut borsoni (110,000kg), and the Asian straight-tusked elephant (~220,000kg), the largest known land mammal [@Larramendi:20151c2]. Remarkably, these large-bodied _Afrotherian_ lineages are nested within small bodied species (__Figure \@ref(fig:Figure-intro)__) [@OLeary2013a; @Springer2013; @OLeary2013b; @PuttickAndThomas2015], indicating that gigantism independently evolved in hyraxes, sea cows, and elephants (_Paenungulata_). Thus, Paenungulates are an excellent model system in which to explore the mechanisms that underlie the evolution of large body sizes and augmented cancer resistance.

Although many mechanisms can potentially resolve Peto’s paradox, among the most parsimonious routes to enhanced cancer resistance is through an increased copy number of tumor suppressors. Indeed, candidate genes studies have found that the elephant genome encodes duplicate such as _TP53_ and _LIF_ [@Abegglen:JAMA2015; @Sulak2015; @Vazquez2018] as well as other genes with putative tumor suppressive functions [@Caulin2015; @Doherty2016]. As these studies focus on _a priori_ gene sets, however, it remains unknown whether this is a general, genome-wide trend in _Afrotherian_ genomes; and whether such a general trend is associated with the recent increases in body size – and therefore expected cancer risk – in these species.

Here, we trace the evolution of body mass and gene copy number variation in across _Afrotherian_ genomes in order to investigate whether duplication of tumors suppresor is common in large, long-lived Proboscideans. Our estimates of the evolution of body mass, similarly to previous studies [@OLeary2013a; @Springer2013; @OLeary2013b; @PuttickAndThomas2015], show that large body masses evolved in a step-wise manner, with major increases in body mass in the _Pseudoungulata_ (17kg), _Paenungulata_ (25kg), _Tethytheria_ (296kg), and _Proboscidea_ (4,100kg) stem-lineages. To explore whether duplication of tumor suppressor genes occurred coincident with the evolution of large body sizes, we used a genome-wide Reciprocal Best BLAT Hit (RBBH) strategy to identify gene duplications, and used maximum likelihood to infer the lineages in which those duplications occurred. Unexpectedly, we found that duplication of tumor suppressor genes was common in all Afrotherians, both large and small. These data suggest that duplication of tumor suppressor genes is pervasive in Afrotherians and proceeded the evolution of species with very large body sizes.

@@

We built a time-calibrated supertree of Eutherian mammals by combining the time-calibrated molecular phylogeny of Bininda-Emonds _et al._ [-@Bininda-Emonds2008] with the time-calibrated total evidence Afrotherian phylogeny from Puttick and Thomas [@PuttickAndThomas2015]. While the Bininda-Emonds _et al._ [-@Bininda-Emonds2008] phylogeny includes 1,679 species, only 34 are Afrotherian, and no fossil data are included. The inclusion of fossil data from extinct species is essential to ensure that ancestral state reconstructions of body mass are not biased by only including extant species. This can lead to inaccurate reconstructions, for example, if lineages convergently evolved large body masses from a small bodied ancestor. In contrast, the total evidence Afrotherian phylogeny of Puttick and Thomas [-@PuttickAndThomas2015] includes 77 extant species and fossil data from 39 extinct species. Therefore we replaced the Afrotherian clade in the Bininda-Emonds _et al._ [-@Bininda-Emonds2008] phylogeny with the Afrotherian phylogeny of Puttick and Thomas [-@PuttickAndThomas2015] using Mesquite. Next, we jointly estimated rates of body mass evolution and reconstructed ancestral states using a generalization of the Brownian motion model that relaxes assumptions of neutrality and gradualism by considering increments to evolving characters to be drawn from a heavy-tailed stable distribution (the "Stable Model") implemented in StableTraits [@ElliotAndMooers2014]. The stable model allows for occasional large jumps in traits and has previously been shown to out-perform other models of body mass evolution, including standard Brownian motion models, Ornstein–Uhlenbeck models, early burst maximum likelihood models, and heterogeneous multi-rate models [@ElliotAndMooers2014].

@@

__Reciprocal Best-Hit BLAT:__ We developed a reciprocal best hit BLAT (RBHB) pipeline to identify putative homologs and estimate gene copy number across species (\@ref(fig:Figure-RBHBStrategy)). The Reciprocal Best Hit (RBH) search strategy is conceptually straightforward: 1) Given a gene of interest $G_A$ in a query genome $A$, one searches a target genome $B$ for all possible matches to $G_A$; 2) For each of these hits, one then performs the reciprocal search in the original query genome to identify the highest-scoring hit; 3) A hit in genome $B$ is defined as a homolog of gene $G_A$ if and only if the original gene $G_A$ is the top reciprocal search hit in genome $A$. We selected BLAT [@blat] as our algorithm of choice, as this algorithm is sensitive to highly simliar (>90% identity) sequences, thus identifying the highest-confidence homologs while minimizing many-to-one mapping problems when searching for multiple genes. RBH performs similar to other more complex methods of orthology prediction, and is particularly good at identifying incomplete genes that may be fragmented in low quality/poor assembled regions of the genome [@AltenhoffAndDessimoz2009; @SalichosAndRokas2011].

__Effective Copy Number By Coverage:__ In lower-quality genomes, many genes are fragmented across multiple scaffolds, which results in BLAT calling multiple hits when in reality there is only one gene. To compensate for this, we came up with a novel statistic, Estimated Copy Number by Coverage (ECNC), which averages the number of times we see each nucleotides of a query sequence in a target genome over the total number of nucleotides of the query sequence found overall in each target genome (__Figure \@ref(fig:Figure-ECNC)__). This allows us to correct for genes that have been fragmented across incomplete genomes, while also taking into account missing sequences from the human query in the target genome. Mathematically, this can be written as:  
$$ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}$$
where $n$ is a given nucleotide in the query, $l$ is the total length of the query, $C_n$ is the number of instances that $n$ is present within a reciprocal best hit, and $bool(C_n)$ is 1 if $C_n > 0$ or 0 if $C_n = 0$.  
```{r RBHB Stats}
n.TranscriptsSearched <- query.binTable %>% filter(!C.ORF, !LOC, !LINC, !HLA, !Histone, !ZincFinger,!Riboprotein, !OlfactoryReceptor, !Uncharacterized, !Putative, !Viral, !Fragment) %>% pull(ID) %>% unique %>% length()
n.GenesSearched <- query.binTable %>% filter(!C.ORF, !LOC, !LINC, !HLA, !Histone, !ZincFinger,!Riboprotein, !OlfactoryReceptor, !Uncharacterized, !Putative, !Viral, !Fragment) %>% pull(Gene) %>% unique %>% length()
genesPerGenome <- geneCopyTable %>% filter(Copy>=1, !is.na(ECNC)) %>% group_by(Genome) %>% summarise(
  Hits=n(), pc.Found=Hits/n.GenesSearched, 
  mean.ECNC = mean(ECNC), q1.ECNC = quantile(ECNC, 1/4), median.ECNC = median(ECNC), q3.ECNC = quantile(ECNC, 3/4), max.ECNC = max(ECNC), sd.ECNC = sd(ECNC), 
  mean.Copy = mean(Copy), q1.Copy = quantile(Copy, 1/4), median.Copy = median(Copy), q3.Copy = quantile(Copy, 3/4), max.Copy = max(Copy), sd.Copy = sd(Copy), 
  mean.ECNCPerHit = mean(ECNC/Copy), q1.ECNCPerHit = quantile(ECNC/Copy, 1/4), median.ECNCPerHit = median(ECNC/Copy), q3.ECNCPerHit = quantile(ECNC/Copy, 3/4), max.ECNCPerHit = max(ECNC/Copy), min.ECNCPerHit = min(ECNC/Copy), sd.ECNCPerHit = sd(ECNC/Copy))
duplicatedGenesPerGenome <- geneCopyTable %>% filter(Copy>=1, ECNC>=1.5) %>% group_by(Genome) %>% summarise(Duplicated=n())

@@

After filtering out problematic protein queries (see below), we then used our pipeline to search for all copies of our `r n.GenesSearched` query genes in publicly available _Afrotherian_ genomes (\@ref(tab:Table-Genomes)), including African savannah elephant (_Loxodonta africana_: loxAfr3, loxAfr4, loxAfrC), African forest elephant (_Loxodonta cyclotis_: loxCycF), Asian Elephant (_Elephas maximus_: eleMaxD), Woolly Mammoth (_Mammuthus primigenius_: mamPriV), Colombian mammoth (_Mammuthus columbi_: mamColU), American mastodon (_Mammut americanum_: mamAmeI), Rock Hyrax (_Procavia capensis_: proCap1, proCap2, proCap2_HiC), West Indian Manatee (_Trichechus manatus latirostris_: triManLat1, triManLat1_HiC), Aardvark (_Orycteropus afer_: oryAfe1, oryAfe1_HiC), Lesser Hedgehog Tenrec (_Echinops telfairi_: echTel2), Nine-banded armadillo (_Dasypus novemcinctus_: dasNov3), Hoffman's two-toed sloth (_Choloepus hoffmannii_: choHof1, choHof2, choHof2_HiC), Cape golden mole (_Chrysochloris asiatica_: chrAsi1), and Cape elephant shrew (_Elephantulus edwardii_: eleEdw1).

__Query gene inclusion criteria:__ To assemble our query list, we first removed all unnamed genes from UP000005640. Next, we excluded genes from downstream analyses for which assignment of homology was uncertain, including uncharacterized ORFs (`r query.binTable %>% filter(C.ORF) %>% nrow` genes), LOC (`r query.binTable %>% filter(LOC) %>% nrow` genes), HLA genes (`r query.binTable %>% filter(HLA) %>% nrow` genes), replication dependent histones (`r query.binTable %>% filter(Histone) %>% nrow` genes), odorant receptors (`r query.binTable %>% filter(OlfactoryReceptor) %>% nrow` genes), ribosomal proteins (`r query.binTable %>% filter(Riboprotein) %>% nrow` genes), zinc finger transcription factors (`r query.binTable %>% filter(ZincFinger) %>% nrow` genes), viral and repetitive-element-associated proteins (`r query.binTable %>% filter(Viral) %>% nrow` genes) and any protein described as either "Uncharacterized," "Putative," or "Fragment" by UniProt in UP000005640 (`r query.binTable %>% filter(Fragment | Uncharacterized | Putative) %>% nrow` genes), leaving us with a final set of `r n.TranscriptsSearched` query protein isoforms, corresponding to `r n.GenesSearched` genes.

__Duplication gene inclusion criteria:__ In order to condense transcript-level hits into single gene loci, and to resolve many-to-one genome mappings, we removed exons where transcripts from different genes overlapped, and merged overlapping transcripts of the same gene into a single gene locus call. The resulting gene-level copy number table was then combined with the maximum ECNC values observed for each gene in order to call gene duplications. We called a gene duplicated if its copy number was two or more, and if the maximum ECNC value of all the gene transcripts searched was 1.5 or greater; previous studies have shown that incomplete duplications can encode functional genes [@Sulak2015; @Vazquez2018], therefore partial gene duplications were included provided they passed additional inclusion criteria (see below). The ECNC cut off of 1.5 was selected empirically, as this value minimized the number of false positives seen in a test set of genes and genomes. The results of our initial search are summarized in Figure \@ref(fig:Figure-GeneDup-Cladogram). Overall, we identified `r median.genesPerGenome` genes across all species, or `r median.genesPerGenome %>% divide_by(n.GenesSearched) %>% multiply_by(100) %>% round(digit=1) %>% paste0(.,"%")` of our starting query genes.

@@

## Evidence for Functionality of Gene Duplicates

@@

__Transcriptome Assembly:__ For the African Savana Elephant, Asian Elephant, West Indian Manatee, and Nine-Banded Armadillo, we generated _de novo_ transcriptomes using publically-available RNA-sequencing data from NCBI SRA (\@ref(tab:Table-SRAData)). We mapped reads to all genomes available for each species, and assembled transcripts using HISAT2 and StringTie, respectively [@HISAT; @StringTie; @Tuxedo]. RNA-sequencing data was not available for Cape Golden Mole, Cape Elephant Shrew, Rock Hyrax, Aardvark, or the Lesser Hedgehog Tenrec.

@@

To determine if gene duplications were enriched in particular biological pathways, we used the WEB-based Gene SeT AnaLysis Toolkit (WEBGESTALT)[@WebGestalt2019] to perform overrepresentation analysis (ORA); pathway databases included Reactome [@Reactome], Wikipathways, [@Wikipathways], and KEGG [@KEGG]. Gene duplicates in each lineage were used as the foreground gene set, and the initial query set was used as the background gene set. Statistical significance of enriched terms was assessed with a hypergeometric test, and controlled for multiple testing using the Benjamini-Hochberg false discovery rate (FDR); each analysis was run at FDR=0.1, FDR=0.2, FDR=0.3, and FDR=0.5. In order to determine an empirical false positive rate for term enrichment, we randomly sampled 1000-10,000 genes from our background set 1000 times, and ran the aforementioned analyses to see which terms were likely to randomly appear; there were no terms which appeared at FDR $\leq$ 0.3.

## Estimating the Evolution of Cancer Risk
